---
title: "Replications of RQ2-4"
output:
  pdf_document: default
params:
  data:  
    input: file  
    label: 'LIWC 2007 dataset:'  
    value: datasets/pers_liwc07_nlon.csv 
  commits:
    input: file
    label: 'Devs commit activities:'
    value: datasets/commits.csv 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn=1)
options(digits=3)
set.seed(1977)
library(readr)
library(zoo)
library(grid)
library(gridExtra)
library(effsize)
library(metan)
library(sqldf)
library(ggplot2)
```

## Data preparation 
We load the file with the scores from LIWC and rescale them in the range [1,5].
```{r load scores, message=FALSE, warning=FALSE, include=TRUE}
personality = read_delim(params$data, ";", escape_double = FALSE)
personality$openness <- resca(personality,openness, 
                              new_min=1, new_max=5)$openness_res 
personality$conscientiousness <- resca(personality,conscientiousness, 
                                       new_min=1, new_max=5)$conscientiousness_res  
personality$extraversion <- resca(personality,extraversion, 
                                  new_min=1, new_max=5)$extraversion_res  
personality$agreeableness <- resca(personality,agreeableness, 
                                   new_min=1, new_max=5)$agreeableness_res  
personality$neuroticism <- resca(personality,neuroticism, 
                                 new_min=1, new_max=5)$neuroticism_res
```

Now we load the commit data and merge them with the personality data.
```{r load commits, message=FALSE, warning=FALSE, include=TRUE}
commit = read_delim(params$commits, ";", escape_double = FALSE)

# Find which developers appear in the intersection of the two data sets
both = intersect(unique(commit$uid), unique(personality$uid))

# Extract data only for the intersection developers
# (filter out people with 0 commits)
commit.both = subset(commit, uid %in% both & 
                       num_authored_commits > 0)
# Count total number of projects and total number of commits across all projects, per person
commit.count = sqldf("select uid, 
                     sum(num_authored_commits) as 'total_commits', 
                     count(project) as 'total_projects',
                     max(last_authored_datetime) as last_authored_datetime,
                     min(first_integrated_datetime) as first_integrated_datetime
                     from `commit.both` group by uid")

# Filter out people who are still active (have at least one commit 
# during the last 3 months before data collection). We can't know
# if they will remain one-time contributors or have more commits
commit.count = subset(commit.count, 
                      last_authored_datetime < as.POSIXct("2017-09-01 20:18:02"))

# Identify people with only one commit total, across all projects.
# These are the one-timers. The others are more active, even if they
# have projects with only one commit
one.timers = subset(commit.count, total_commits == 1)$uid
multi.timers = subset(commit.count, total_commits > 1)$uid
one.or.multi.timers = subset(commit.count, total_commits >= 1)$uid

# Assign a binary label "one_timer" to everyone in the personality 
# data, based on the distinction above
p = subset(personality, uid %in% one.timers | uid %in% multi.timers)
p$one_timer = FALSE
for (i in 1:nrow(p)){
  p[i,]$one_timer = p[i,]$uid %in% one.timers
}

# Compute average personality scores per person, across time and across all projects
# We will use this simple data in the analysis below, because personality doesn't
# change much (see boxplots)
p.aggr = sqldf("select uid, avg(openness) as 'openness', 
               avg(agreeableness) as 'agreeableness',
               avg(neuroticism) as 'neuroticism',
               avg(extraversion) as 'extraversion', 
               avg(conscientiousness) as 'conscientiousness',
               sum(word_count) as word_count
               from p group by uid")

# Apply the "one_timer" binary label to this aggregate data set
p.aggr$one_timer = FALSE
for (i in 1:nrow(p.aggr)){
  p.aggr[i,]$one_timer = p.aggr[i,]$uid %in% one.timers
}
# Apply the "multi_timer" binary label to this aggregate data set
p.aggr$m_timer = FALSE
for (i in 1:nrow(p.aggr)){
  p.aggr[i,]$m_timer = p.aggr[i,]$uid %in% multi.timers
}
```

## RQ2 Personality of core vs. peripheral developers

We perform a series o Wilcoxon Mann Whitney for unpaired groups, one for each trait. We adjust the significance with Bonferroni due to multiple comparisons. The differences are not significant (adjuste p-values all > 0.05).

```{r core vs peripheral}
integrations = subset(commit, uid %in% both & num_integrated_commits > 0)
integration.count = sqldf("select uid, 
                          min(first_integrated_datetime) as first_integrated_datetime,
                          max(last_integrated_datetime) as last_integrated_datetime, 
                          num_integrated_commits
                          from `integrations` group by uid")
integration.count = subset(integration.count, last_integrated_datetime < as.POSIXct("2017-09-01 20:18:02"))
integrators = subset(integration.count, num_integrated_commits >= 1)$uid

# peripheral
authors_non_integrators = unique(commit.both[!commit.both$uid %in% integrators, ]$uid)
c_b_n_i = Reduce(intersect, list(commit.both$uid, authors_non_integrators))
peripheral = personality[personality$uid %in% c_b_n_i, ]
peripheral = sqldf::sqldf("select uid, 
                          avg(openness) as openness,
                          avg(conscientiousness) as conscientiousness,
                          avg(extraversion) as extraversion,
                          avg(agreeableness) as agreeableness,
                          avg(neuroticism) as neuroticism
                          from peripheral
                          group by uid")
# core
c_b_i = Reduce(intersect, list(commit.both$uid, integrators))
core = personality[personality$uid %in% c_b_i, ]
core = sqldf::sqldf("select uid, 
                          avg(openness) as openness,
                          avg(conscientiousness) as conscientiousness,
                          avg(extraversion) as extraversion,
                          avg(agreeableness) as agreeableness,
                          avg(neuroticism) as neuroticism
                          from core
                          group by uid")

print("**************************")
print("          openness       ")
print("**************************")
wto = wilcox.test(peripheral$openness, core$openness, conf.int=TRUE, paired=FALSE)
wto
#cliff.delta(peripheral$openness, core$openness, return.dm=TRUE)

print("**************************")
print("    conscientiousness    ")
print("**************************")
wtc = wilcox.test(peripheral$conscientiousness, core$conscientiousness, conf.int=TRUE, paired=FALSE)
wtc
#cliff.delta(peripheral$conscientiousness, core$conscientiousness, return.dm=TRUE)

print("**************************")
print("       extraversion      ")
print("**************************")
wte = wilcox.test(peripheral$extraversion, core$extraversion, conf.int=TRUE, paired=FALSE)
wte
#cliff.delta(peripheral$extraversion, core$extraversion, return.dm=TRUE)

print("**************************")
print("       agreeableness     ")
print("**************************")
wta = wilcox.test(peripheral$agreeableness, core$agreeableness, conf.int=TRUE, paired=FALSE)
wta
#cliff.delta(peripheral$agreeableness, core$agreeableness, return.dm=TRUE)

print("**************************")
print("        neuroticism      ")
print("**************************")
wtn = wilcox.test(peripheral$neuroticism, core$neuroticism, conf.int=TRUE, paired=FALSE)
wtn
#cliff.delta(peripheral$neuroticism, core$neuroticism, return.dm=TRUE)

print("Bonferroni")
p.adjust(c(wto$p.value, wtc$p.value, wte$p.value, wta$p.value, wtn$p.value), method = "bonferroni")

```

# RQ3 Individual personality differences before and after becoming a project contributor
We use a series Wilcoxon Signed-Rank test for paired samples (before and after), repeated for each of the five traits.
All the test returns differences taht are not statistically significant after Bonferroni corrections for multiple comparisons, as can also been observed from the box plots below.

```{r rq3 preparation, message=FALSE, warning=FALSE, include=TRUE}
df = data.frame(when=character(0), value=integer(0), stringsAsFactors = FALSE)

for (i in 1:length(integrators)){
  split_index = 
    as.Date(integration.count[integration.count$uid == integrators[i], ]$first_integrated_datetime, 
            format="%Y-%m")
  p_tot = personality[personality$uid == integrators[i],]
  p_prev = 
    personality[personality$uid == integrators[i] & as.yearmon(personality$month) < as.yearmon(split_index), ]
  p_aft = 
    personality[personality$uid == integrators[i] & as.yearmon(personality$month) >= as.yearmon(split_index), ]
  
  df[nrow(df)+1, ] = list("O before", mean(p_prev$openness))
  df[nrow(df)+1, ] = list("O after", mean(p_aft$openness))
  df[nrow(df)+1, ] = list("C before", mean(p_prev$conscientiousness))
  df[nrow(df)+1, ] = list("C after", mean(p_aft$conscientiousness))
  df[nrow(df)+1, ] = list("E before", mean(p_prev$extraversion))
  df[nrow(df)+1, ] = list("E after", mean(p_aft$extraversion))
  df[nrow(df)+1, ] = list("A before", mean(p_prev$agreeableness))
  df[nrow(df)+1, ] = list("A after", mean(p_aft$agreeableness))
  df[nrow(df)+1, ] = list("N before", mean(p_prev$neuroticism))
  df[nrow(df)+1, ] = list("N after", mean(p_aft$neuroticism))
}


df_o = df[df$when=="O before" | df$when=="O after", ]
df_o$when <- factor(df_o$when, levels = c("O before", "O after") )
ob = wilcox.test(df[df$when=="O before", ]$value, df[df$when=="O after", ]$value, 
                 correct = FALSE, paired = TRUE, conf.int=TRUE)
ob

df_c = df[df$when=="C before" | df$when=="C after", ]
df_c$when <- factor(df_c$when, levels = c("C before", "C after") )
cb = wilcox.test(df[df$when=="C before", ]$value, df[df$when=="C after", ]$value, 
                 correct = FALSE, paired = TRUE, conf.int=TRUE)
cb

df_e = df[df$when=="E before" | df$when=="E after", ]
df_e$when <- factor(df_e$when, levels = c("E before", "E after") )
eb = wilcox.test(df[df$when=="E before", ]$value, df[df$when=="E after", ]$value, 
                 correct = FALSE, paired = TRUE, conf.int=TRUE)
eb

df_a = df[df$when=="A before" | df$when=="A after", ]
df_a$when <- factor(df_a$when, levels = c("A before", "A after") )
ab=wilcox.test(df[df$when=="A before", ]$value, df[df$when=="A after", ]$value, 
               correct = FALSE,  paired = TRUE, conf.int=TRUE)
ab

df_n = df[df$when=="N before" | df$when=="N after", ]
df_n$when <- factor(df_n$when, levels = c("N before", "N after") )
nb = wilcox.test(df[df$when=="N before", ]$value, df[df$when=="N after", ]$value, 
                 correct = FALSE,  paired = TRUE, conf.int=TRUE)
nb

p.adjust(c(ob$p.value, cb$p.value, eb$p.value, ab$p.value, nb$p.value), method = "bonferroni")
```
```{r boxplots rq3, message=FALSE, warning=FALSE, echo=FALSE}
bplot_o <- ggplot(df_o, aes(x = when, y = value)) +
  geom_boxplot(colour = "black", fill = "white",
               size = 0.4, notch = TRUE) +
  scale_y_continuous(name = "Personality trait score", 
                     limits=c(1, 5)) +
  scale_x_discrete(name = "") +
  ggtitle("Openness") +
  theme_bw() +
  stat_boxplot(geom = "errorbar", width = 0.3) +  
  theme(panel.grid.major = element_blank(),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        text=element_text(),
        axis.title = element_text(face="bold"),
        axis.text.x=element_text(colour="black", size = 11),
        axis.text.y=element_text(colour="black", size = 9),
        axis.line = element_line(size=0.5, colour = "black"))


bplot_c <-ggplot(df_c, aes(x = when, y = value)) +
  geom_boxplot(colour = "black", fill = "white",
               size = 0.4, notch = TRUE) +
  scale_y_continuous(name = "", 
                     limits=c(1, 5)) +
  scale_x_discrete(name = "") +
  ggtitle("Conscientious.") +
  theme_bw() +
  stat_boxplot(geom = "errorbar", width = 0.3) +  
  theme(panel.grid.major = element_blank(),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        text=element_text(),
        axis.title = element_text(face="bold"),
        axis.text.x=element_text(colour="black", size = 11),
        axis.text.y=element_text(colour="black", size = 9),
        axis.line = element_line(size=0.5, colour = "black"))


bplot_e <-ggplot(df_e, aes(x = when, y = value)) +
  geom_boxplot(colour = "black", fill = "white",
               size = 0.4, notch = TRUE) +
  scale_y_continuous(name = "", 
                     #breaks = seq(0, 1, 25),
                     limits=c(1, 5)) +
  scale_x_discrete(name = "") +
  ggtitle("Extraversion") +
  theme_bw() +
  stat_boxplot(geom = "errorbar", width = 0.3) +  
  theme(panel.grid.major = element_blank(),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        text=element_text(),
        axis.title = element_text(face="bold"),
        axis.text.x=element_text(colour="black", size = 11),
        axis.text.y=element_text(colour="black", size = 9),
        axis.line = element_line(size=0.5, colour = "black"))


bplot_a <-ggplot(df_a, aes(x = when, y = value)) +
  geom_boxplot(colour = "black", fill = "white",
               size = 0.4, notch = TRUE) +
  scale_y_continuous(name = "Personality trait score", 
                     #breaks = seq(0, 1, 25),
                     limits=c(1, 5)) +
  scale_x_discrete(name = "") +
  ggtitle("Agreeableness") +
  theme_bw() +
  stat_boxplot(geom = "errorbar", width = 0.3) +  
  theme(panel.grid.major = element_blank(),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        text=element_text(),
        axis.title = element_text(face="bold"),
        axis.text.x=element_text(colour="black", size = 11),
        axis.text.y=element_text(colour="black", size = 9),
        axis.line = element_line(size=0.5, colour = "black"))


bplot_n <-ggplot(df_n, aes(x = when, y = value)) +
  geom_boxplot(colour = "black", fill = "white",
               size = 0.4, notch = TRUE) +
  scale_y_continuous(name = "", 
                     limits=c(1, 5)) +
  scale_x_discrete(name = "") +
  ggtitle("Neuroticism") +
  theme_bw() +
  stat_boxplot(geom = "errorbar", width = 0.3) +  
  theme(panel.grid.major = element_blank(),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        text=element_text(),
        axis.title = element_text(face="bold"),
        axis.text.x=element_text(colour="black", size = 11),
        axis.text.y=element_text(colour="black", size = 9),
        axis.line = element_line(size=0.5, colour = "black"))
grid.arrange(bplot_o, bplot_c, bplot_e, bplot_a, bplot_n, ncol = 3, nrow=2)
```


## RQ4 — variation with the degree of development activity

Starting from the core and peripheral groups of developers of RQ2, we further split the groups into high vs. low contributors.
We then perform the unpaired comparisons of the median personality scores between these high vs. low-activity developers within the core and periphearl groups. Results are in shown in the table below. The Wilcoxon Rank Sum tests reveal no cases of statistically significant differences between the pairs of trait distributions (i.e., adjusted p-values > 0.05 after Bonferroni correction).

```{r}
csbo = sqldf::sqldf("select uid, 
                         sum(num_authored_commits) as num_authored_commits,
                         sum(num_integrated_commits) as num_integrated_commits
                         from `commit.both`
                         group by uid")
# split peripheral in low vs higher commit authors around the mean
Mp = floor(mean(csbo$num_authored_commits))
peripheral_low = fn$sqldf("select *
                              from peripheral left join csbo on peripheral.uid = csbo.uid
                              where num_authored_commits <= $Mp")
peripheral_high = fn$sqldf("select *
                              from peripheral left join csbo on peripheral.uid = csbo.uid
                              where num_authored_commits > $Mp")


wtc_hlpo = wilcox.test(peripheral_high$openness, peripheral_low$openness, 
                       conf.int=TRUE, paired=FALSE)
#cliff.delta(peripheral_high$openness, peripheral_low$openness, return.dm=TRUE)

wtc_hlpc = wilcox.test(peripheral_high$conscientiousness, peripheral_low$conscientiousness, 
                       conf.int=TRUE, paired=FALSE)
#cliff.delta(peripheral_high$conscientiousness, peripheral_low$conscientiousness, return.dm=TRUE)

wtc_hlpe = wilcox.test(peripheral_high$extraversion, peripheral_low$extraversion, 
                       conf.int=TRUE, paired=FALSE)
#cliff.delta(peripheral_high$extraversion, peripheral_low$extraversion, return.dm=TRUE)

wtc_hlpa = wilcox.test(peripheral_high$agreeableness, peripheral_low$agreeableness, 
                       conf.int=TRUE, paired=FALSE)
#cliff.delta(peripheral_high$agreeableness, peripheral_low$agreeableness, return.dm=TRUE)

wtc_hlpn = wilcox.test(peripheral_high$neuroticism, peripheral_low$neuroticism, 
                       conf.int=TRUE, paired=FALSE)
#cliff.delta(peripheral_high$neuroticism, peripheral_low$neuroticism, return.dm=TRUE)

adj_p <- p.adjust(c(wtc_hlpo$p.value, wtc_hlpc$p.value, wtc_hlpe$p.value, wtc_hlpa$p.value, wtc_hlpn$p.value), 
                  method = "bonferroni")


df <- data.frame(matrix(ncol = 5, nrow = 12))
x <- c("Trait", "W", "p-value ", "CI 95% low", "CI 95% high")
colnames(df) <- x
df[1,] <- c("High vs. low commit authors (peripheral)", "", "", "", "")
df[2,] <- c("openness", as.numeric(wtc_hlpo$statistic), adj_p[1], round(as.numeric(wtc_hlpo$conf.int), 3))
df[3,] <- c("conscientiousness", as.numeric(wtc_hlpc$statistic), adj_p[2], round(as.numeric(wtc_hlpc$conf.int), 3))
df[4,] <- c("extraversion", as.numeric(wtc_hlpo$statistic), adj_p[3], round(as.numeric(wtc_hlpe$conf.int), 3))
df[5,] <- c("agreeableness", as.numeric(wtc_hlpa$statistic), adj_p[4], round(as.numeric(wtc_hlpa$conf.int), 3))
df[6,] <- c("neuroticism", as.numeric(wtc_hlpn$statistic), adj_p[5], round(as.numeric(wtc_hlpn$conf.int), 3))


# split core in low vs higher commit integrators around the mean
Mc = ceiling(mean(csbo$num_integrated_commits))
core_low = fn$sqldf("select *
                        from core left join csbo on core.uid = csbo.uid
                        where num_authored_commits <= $Mc")
core_high = fn$sqldf("select *
                        from core left join csbo on core.uid = csbo.uid
                        where num_authored_commits > $Mc")

wtc_hlco = wilcox.test(core_high$openness, core_low$openness, 
                       conf.int=TRUE, paired=FALSE)
#cliff.delta(core_high$openness, core_low$openness, return.dm=TRUE)

wtc_hlcc = wilcox.test(core_high$conscientiousness, core_low$conscientiousness, 
                       conf.int=TRUE, paired=FALSE)
#cliff.delta(core_high$conscientiousness, core_low$conscientiousness, return.dm=TRUE)

wtc_hlce = wilcox.test(core_high$extraversion, core_low$extraversion, 
                       conf.int=TRUE, paired=FALSE)
#cliff.delta(core_high$extraversion, core_low$extraversion, return.dm=TRUE)

wtc_hlca = wilcox.test(core_high$agreeableness, core_low$agreeableness, 
                       conf.int=TRUE, paired=FALSE)
#cliff.delta(core_high$agreeableness, core_low$agreeableness, return.dm=TRUE)

wtc_hlcn = wilcox.test(core_high$neuroticism, core_low$neuroticism, 
                       conf.int=TRUE, paired=FALSE)
#cliff.delta(core_high$neuroticism, core_low$neuroticism, return.dm=TRUE)

adj_p<-p.adjust(c(wtc_hlco$p.value, wtc_hlcc$p.value, wtc_hlce$p.value, wtc_hlca$p.value, wtc_hlcn$p.value), 
                method = "bonferroni") 

df[7,] <- c("High vs. low commit authors (core)", "", "", "", "")
df[8,] <- c("openness", as.numeric(wtc_hlco$statistic), adj_p[1], round(as.numeric(wtc_hlco$conf.int), 3))
df[9,] <- c("conscientiousness", as.numeric(wtc_hlcc$statistic), adj_p[2], round(as.numeric(wtc_hlcc$conf.int), 3))
df[10,] <- c("extraversion", as.numeric(wtc_hlce$statistic), adj_p[3], round(as.numeric(wtc_hlce$conf.int), 3))
df[11,] <- c("agreeableness", as.numeric(wtc_hlca$statistic), adj_p[4], round(as.numeric(wtc_hlca$conf.int), 3))
df[12,] <- c("neuroticism", as.numeric(wtc_hlcn$statistic), adj_p[5], round(as.numeric(wtc_hlcn$conf.int), 3))

df
```

