---
title: "Replication of RQ5-6 regression analyses"
output: pdf_document
params:
  data:  
    input: file  
    label: 'LIWC 2007 dataset:'  
    value: datasets/pers_liwc07_nlon.csv  
  commits:
    input: file
    label: 'Devs commit activities:'
    value: datasets/commits.csv
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1977)
options(digits=3)
```
```{r load libraries, message=FALSE, warning=FALSE, include=FALSE}
library(readr)
library(sqldf)
library(car)
library(arm)
library(pscl)
library(ROCR)
library(caret)
library(lmtest)
library(MASS)
library(metan)
```

## Data preparation 
We load the file with the scores from LIWC and rescale them in the range [1,5].
```{r load scores, message=FALSE, warning=FALSE}
personality = read_delim(params$data, ";", escape_double = FALSE)
personality$openness <- resca(personality,openness, 
                              new_min=1, new_max=5)$openness_res 
personality$conscientiousness <- resca(personality,conscientiousness, 
                                       new_min=1, new_max=5)$conscientiousness_res  
personality$extraversion <- resca(personality,extraversion, 
                                  new_min=1, new_max=5)$extraversion_res  
personality$agreeableness <- resca(personality,agreeableness, 
                                   new_min=1, new_max=5)$agreeableness_res  
personality$neuroticism <- resca(personality,neuroticism, 
                                 new_min=1, new_max=5)$neuroticism_res
```

Now we load the commit data and merge them with the personality data.
```{r load commits, message=FALSE, warning=FALSE}
commit = read_delim(params$commits, ";", escape_double = FALSE)
# Find which developers appear in the intersection of the two data sets
both = intersect(unique(commit$uid), unique(personality$uid))
# Extract data only for the intersection developers
# (filter out people with 0 commits)
commit_both = subset(commit, uid %in% both &
                       num_authored_commits > 0)
id_no_cotributors = setdiff(unique(personality$uid), unique(commit$uid))
pers_zero_contributors = subset(personality, uid %in% id_no_cotributors)

commit.count = sqldf(
  "select uid, sum(num_authored_commits) as 'total_commits',
  count(project) as 'total_projects',
  max(last_authored_datetime) as last_authored_datetime,
  min(first_integrated_datetime) as first_integrated_datetime
  from `commit_both` group by uid"
)

# Filter out people who are still active (have at least one commit
# during the last 3 months before data collection). We can't know
# if they will remain one-time contributors or have more commits
commit.count = subset(commit.count,
                      last_authored_datetime < as.POSIXct("2017-09-01 20:18:02"))

# Identify people with only one commit total, across all projects.
# These are the one-timers. The others are more active, even if they
# have projects with only one commit
one.timers = subset(commit.count, total_commits == 1)$uid
multi.timers = subset(commit.count, total_commits > 1)$uid
oneormore.timers = subset(commit.count, total_commits >= 1)$uid

# Assign a binary label "one_timer" to everyone in the personality
# data, based on the distinction above
p = subset(personality, uid %in% one.timers | uid %in% multi.timers)
p$one_timer = FALSE
for (i in 1:nrow(p)) {
  p[i,]$one_timer = p[i,]$uid %in% one.timers
}

p = subset(personality, uid %in% one.timers | uid %in% multi.timers)

# Compute average personality scores per person, across time and across all projects
p.aggr = sqldf(
  "select uid, avg(openness) as 'openness', 
  avg(agreeableness) as 'agreeableness',
  avg(neuroticism) as 'neuroticism', 
  avg(extraversion) as 'extraversion',
  avg(conscientiousness) as 'conscientiousness',
  sum(word_count) as word_count,
  project
  from p group by uid"
)

# Apply the "one_timer" label to this aggregate data set
p.aggr$one_timer = FALSE
for (i in 1:nrow(p.aggr)) {
  p.aggr[i,]$one_timer = p.aggr[i,]$uid %in% one.timers
}
# Apply the "mulit_timer" label to this aggregate data set
p.aggr$m_timer = FALSE
for (i in 1:nrow(p.aggr)) {
  p.aggr[i,]$m_timer = p.aggr[i,]$uid %in% multi.timers
}
# calculate project age
proj_age = sqldf::sqldf("select project, project_age 
                        from `commit` 
                        group by project")
# add this pieace of info to the aggregated dataset
p.aggr = p.aggr[p.aggr$project %in% unique(proj_age$project),]
p.aggr$project_age = 0
for (i in 1:nrow(p.aggr)) {
  p.aggr[i,]$project_age = 
    proj_age[proj_age$project == p.aggr[i,]$project,]$project_age
}
```
## RQ5: logistic regression model of contribution likelihood

We build a simple logistic regression model to explain whether someone is a one-time contributor or not based on their personality score, controlling for the number of words in their emails (which may influence personality).
Accordingly, the dependent, predicted variable is whether or not a one-time contributor will make further contributions.

First we analyze potential high correlations in the dataset. We drop *neuroticism* and *extraversion* because they shows high correlation (~70) with *conscientiousness* and *agreeableness*, respectively.


```{r corrplot, echo=FALSE}
# only numerical data
query = sqldf(
  "select openness, 
  agreeableness, 
  neuroticism, 
  extraversion, 
  conscientiousness,
  word_count, 
  project_age
  from `p.aggr`")

matr <- cor(query, method = "pearson",  use = "complete.obs")
corrplot::corrplot(matr, type = "lower", method = "number")

```


```{r glm, echo=TRUE, message=FALSE, warning=FALSE}
m = glm(
  m_timer ~ log(word_count) +
    project_age +
    openness +
    #agreeableness +
    #neuroticism +
    extraversion +
    conscientiousness,
  data = p.aggr,
  family = binomial(link = 'logit')
)
```

The results show that only the control variable *project age* has a significant, negative effect.
```{r glm metrics, warning=FALSE, include=TRUE}
arm::display(m)
summary(m)
car::Anova(m)
DescTools::PseudoR2(m)
```

The VIF index confirm that there are no collinarearity issues as all the scores are < 4.
```{r collinearity}
car::vif(m)
```

Finally, we compute the AUC to assess the goodness of the model at predicting whether a one-time contributor will become a contributor (i.e., will keep on contributing afterwards).

```{r AUC, echo=TRUE, message=FALSE, warning=FALSE}

trainIndex <-
  caret::createDataPartition(p.aggr$m_timer,
                             p = .7,
                             list = FALSE,
                             times = 1)
p.aggr.train <- p.aggr[trainIndex,]
p.aggr.test  <- p.aggr[-trainIndex,]

drops <- c("uid", "one_timer")
p.aggr.train = p.aggr.train[, !(names(p.aggr.train) %in% drops)]
p.aggr.test = p.aggr.test[, !(names(p.aggr.test) %in% drops)]

m1 = stats::glm(
  m_timer ~ log(word_count) +
    openness +
    #agreeableness +
    #neuroticism +
    extraversion +
    conscientiousness,
  data = p.aggr.train,
  family = binomial(link = 'logit')
)
mp <-
  stats::predict.glm(m1,
                     newdata = p.aggr.test,
                     type = "response",
                     se.fit = TRUE)
#mean(mp$fit)
mp1 <- predict(m, newdata = p.aggr.test, type = "response")
mpr <- prediction(mp1, p.aggr.test$m_timer)
mprf <- performance(mpr, measure = "tpr", x.measure = "fpr")
#plot(mprf)

auc <- performance(mpr, measure = "auc")
auc <- auc@y.values[[1]]
auc
```


## RQ6 panel data regression model of contribution intensity

In this count data model, the predicted variable is the number of commit per developer.

We prepare the data accordingly.

```{r rq6 data prep, include=FALSE}
# we ignore projects as the scores per dev id show very little differences across them
# so we aggregate them
p_proj = sqldf(
  "select uid, avg(openness) as 'openness', 
  avg(agreeableness) as 'agreeableness',
  avg(neuroticism) as 'neuroticism', 
  avg(extraversion) as 'extraversion',
  avg(conscientiousness) as 'conscientiousness',
  sum(word_count) as word_count
  from p group by uid"
)

id_commit = sqldf(
  "select uid,
  project as project_name, 
  is_integrator,
  project_size, 
  project_age,
  num_authored_commits as commits_merged, 
  num_integrated_commits,     
  author_track_record_days
  from commit_both"
)

panel_data <- id_commit
panel_data$openness = 0
panel_data$conscientiousness = 0
panel_data$extraversion = 0
panel_data$agreeableness = 0
panel_data$neuroticism = 0

panel_data <- merge(id_commit, p_proj, by = c("uid"), all.x = FALSE)
# use to ensure NAs removal
panel_data <- panel_data[complete.cases(panel_data),]
unlist(lapply(panel_data, function(x)
  any(is.na(x))))
```

We repeat again the correlation analysis. We find again the same high correlations between *extraversion* and *agreebleness* , and *conscientiousness* and *neuroticism*. We keep the former from each couple because they give better VIF scores as shown below.


```{r rq6 corr analysis, echo=FALSE}
matr <-
  panel_data[, names(panel_data) %in% c(
    "openness",
    "conscientiousness",
    "extraversion",
    "agreeableness",
    "neuroticism",
    "project_age",
    "author_track_record_days"
  )]
m <- cor(matr, method = "pearson",  use = "complete.obs")
corrplot::corrplot(m, type = "lower", method = "number")
```

For a correct count data model analysis, we compare two regression strategies, i.e., Poisson and Negative Binomial and choose the better one according to the Likelihood Ratio Test.
```{r LRT, include=TRUE}
mod_poisson <- glm(
  commits_merged ~
      log(word_count)
    + is_integrator
    + (scale(project_age, center=TRUE, scale = TRUE)) 
  + (scale(author_track_record_days, center=TRUE, scale = TRUE))
  + (scale(openness, center=TRUE, scale = TRUE)) 
  + (scale(conscientiousness, center=TRUE, scale = TRUE)) 
  + (scale(extraversion, center=TRUE, scale = TRUE)) 
  #+ (scale(agreeableness, center=TRUE, scale = TRUE)) 
  #+ (scale(neuroticism, center=TRUE, scale = TRUE))
  , data = panel_data,
  family = "poisson"
)

mod_negbin <- glm.nb(
  commits_merged ~
    log(word_count) 
  + is_integrator
  + (scale(project_age, center=TRUE, scale = TRUE)) 
  + (scale(author_track_record_days, center=TRUE, scale = TRUE))
  + (scale(openness, center=TRUE, scale = TRUE)) 
  + (scale(conscientiousness, center=TRUE, scale = TRUE)) 
  + (scale(extraversion, center=TRUE, scale = TRUE)) 
  #+ (scale(agreeableness, center=TRUE, scale = TRUE)) 
  #+ (scale(neuroticism, center=TRUE, scale = TRUE))
  , data = panel_data
)
lmtest::lrtest(mod_poisson, mod_negbin)
```
 We find that Model 2 (Negative Binomial) fits better and has not multi-collinearity issues:
```{r rq6 vif}
car::vif(mod_negbin)
```
Here is a report of the Negative Binomial model fit. We observe that the only predictors related to personality with a significant effect is *conscientiousness* (coefficient=0.123, p<0.05). Hence, the more organized developers make more commits.
Regarding the control variables, we observe that the authors’ track record (i.e.,
the number of days between their first and last successful contribution) has a positive and
significant association (coefficient=0.566) with the number of their merged contributions
(p<0.001). Similarly, we find a positive and significant association between the response
variable and the fact that a developer is a core team member who has integrated external
contributions (coefficient=0.474, p<0.05). Instead, project age has a significant, negative effect
(-0.087, p<0.05), so the longer the project history, the harder it is to make more commits.
However, the model fits the data marginally (Pseudo-R2=0.109) but this was expected as we did not
aim for model completeness but rather at understanding the effects of personality traits.

```{r neg bin model performance, message=FALSE, warning=FALSE, include=TRUE}
summary(mod_negbin)
car::Anova(mod_negbin)
AIC(mod_negbin)
BIC(mod_negbin)
DescTools::PseudoR2(mod_negbin, which = "all")
```

